---
title: "Homework 3 "
output: github_document
date: 10/16/2024 (Due date)
---

```{r setup, include = FALSE}
library(tidyverse)
```

## Problem 1
```{r data import, include = FALSE}
library(p8105.datasets) 
data("ny_noaa") 

ny_noaa #view variable types

#Make sure to comment on structure of data
# variable units from https://p8105.com/dataset_noaa.html
```

**Dataset Description**

The `ny_noaa` dataset, which was uploaded from the _p8105.dataset_ package, contains `r ncol(ny_noaa)` variables and `r nrow(ny_noaa)` observations. Character variables include weather station ID (`id`), `tmax`, and `tmin`. Integer variables include `prcp`, `snow`, and `snwd`. There is also a `date` variable, which is entered in YYYY-MM-DD format. After viewing the variable types in the dataset, I converted `tmax` and `tmin` into doubles. Using `summary()`, we get the following information:

```{r convert variables, include = FALSE}
ny_noaa_df = ny_noaa |>
  mutate(
    tmax = as.double(tmax),
    tmin = as.double(tmin))

summary(ny_noaa_df)
```

* Precipitation (`prcp`) ranges from 0 to 22860 tenths of mm. The mean is 29.82 tenths of mm. 145838 entries have NA values for this variable.
* Snowfall (`snow`) ranges from -13 mm to 10160 mm. The mean is 5 mm. 381221 entries have NA values for this variable.
* Snow depth (`snwd`) ranges from 0 mm to 9195 mm. The mean is 37.3 mm. 591786 entries have NA values for this variable. 
* Maximum temperature (`tmax`) ranges from -389 to 600 tenths of degrees C with a mean of 139.8. 1132358 entries have  NA values for this variable.
* Mininun temperature (`tmin`) ranges from -594 to 600 tenths of degrees C with a mean of 30.3. 1134420 entries have  NA values for this variable.

As you can see by looking at the number of entries with NAs for each of the variables above, missing data is quite an issue in this dataset. 

#### Task 1


```{r split date, include =FALSE}
ny_noaa_updated = ny_noaa_df |>
   separate(date, into = c("year", "month", "day"), sep = "-")
```

The `date` variable is split into 3 different variables: `year`, `month`, and `day` in the dataset `ny_noaa_updated`.

```{r assign units, include = FALSE}
ny_noaa_df = ny_noaa_df |>
  mutate(
    tmax = tmax/10, #changed units of temperature to 째C
    tmin = tmin/10, #changed units of temperature to 째C
    prcp = prcp/10) #changed units of temperature to mm
```

To ensure observations for temperature, precipitation, and snowfall are given in reasonable units, I converted the values in `tmax` and `tmin` to have 째C as the units (instead of tenths of 째C). Similarly, the unit for precipitation (`prcp`) has been converted to just mm. Snowfall is in reasonable units (mm).

```{r snowcount, include = FALSE}
ny_noaa_sf = ny_noaa_df |>
  mutate(snow = as.factor(snow)) |>
  count(snow, sort = TRUE)
```

For snowfall (`snow`), the top 3 most commonly observed values are 0mm (77.39%, n=2008508), 25mm (1.20%, n=31022), and 13mm (0.89%, n=23095). It may be important to note that 14.69% of the observations had NAs (n=381221). 

The relatively high percentage of empty observations for snowfall is due to inconsistent data collection of each weather station. Some of these weather stations may collect a portion of the variables. In general, it makes sense that most of the observations recorded by the New York state weather stations had 0mm for snow since the data is collected from the time period between January 1, 1981 through December 31, 2010 and not just during winter months. 

#### Task 2
Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers? 

#### Task 3
Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year. 


## Problem 2
```{r demo setup, include = FALSE}
demo_df = 
  read_csv("./data/Accelerometer/nhanes_covar.csv", na = c("NA", ".", ""))

#Variables *MIMS are the MIMS values for each minute of a 24-hour day starting at midnight.
```


```{r accelerometer setup, include = FALSE}
accel_df =
  read_csv("./data/Accelerometer/nhanes_accel.csv", na = c("NA", ".", ""))

#data on 250 participants in NHANES study
```

#### Task 1
Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind). 

#### Task 2
Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items. 

#### Task 3
Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot. 

#### Task 4
Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences. 

## Problem 3

```{r CitiBike setup, include = FALSE}
Jan2020_df =
  read_csv("./data/CitiBike/Jan 2020 citi.csv", na = c("NA", ".", ""))

Jan2024_df =
  read_csv("./data/CitiBike/Jan 2024 citi.csv", na = c("NA", ".", ""))

July2020_df =
  read_csv("./data/CitiBike/July 2020 citi.csv", na = c("NA", ".", ""))

July2024_df =
  read_csv("./data/CitiBike/July 2024 citi.csv", na = c("NA", ".", ""))

# data on rides taken on the NYC Citi Bike system. Files contain 1% of all rides with a total duration less than 4 hours in each of four months.
```

#### Task 1

Import, clean, and tidy these data, and describe the resulting dataset. 

#### Task 2

Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members. Comment on these results. 

#### Task 3

Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations. 

#### Task 4
Make a plot to investigate the effects of day of the week, month, and year on median ride duration. This plot can include one or more panels, but should facilitate comparison across all variables of interest. Comment on your observations from this plot. 

#### Task 5

There were relatively few electric Citi Bikes in 2020, but many more are available now. For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration. Comment on your results. 
